---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# charlson

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/ellessenne/charlson?branch=master&svg=true)](https://ci.appveyor.com/project/ellessenne/charlson)
[![Travis-CI Build Status](https://travis-ci.org/ellessenne/charlson.svg?branch=master)](https://travis-ci.org/ellessenne/charlson)
[![Coverage Status](https://img.shields.io/codecov/c/github/ellessenne/charlson/master.svg)](https://codecov.io/github/ellessenne/charlson?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/charlson)](https://cran.r-project.org/package=charlson)

`charlson` is an R package for computing the Charlson comorbidity index using ICD-10 codes.

## Installation

You can install charlson from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ellessenne/charlson")
```

## Example

This is a basic example which shows you how to solve a common problem.

First, we simulate 150 pseudo-ICD-10 diagnostic codes for 10 individuals using the `sample_codes()` function:

```{r simulate-data}
# load charlson
library(charlson)
# set a seed for reproducibility
set.seed(1)
x <- data.frame(
  id = sample(1:10, size = 150, replace = TRUE),
  code = sample_diag(150),
  stringsAsFactors = FALSE)
head(x)
```

Then, we compute the Charlson score, index, and each comorbidity domain:

```{r charlson}
cs = charlson(x = x, id = "id", code = "code")
head(cs)
```

## References

The Charlson comorbidity index computed with this package follows the ICD-10 definition of Quan H, Sundararajan V, Halfon P, Fong A, Burnand B, Luthi JC, et al., _Coding algorithms for definingcComorbidities in ICD-9-CM and ICD-10 administrative data_. Medical Care 2005; 43:1130-1139.