---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# comorbidity

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/ellessenne/comorbidity?branch=master&svg=true)](https://ci.appveyor.com/project/ellessenne/comorbidity)
[![Travis-CI Build Status](https://travis-ci.org/ellessenne/comorbidity.svg?branch=master)](https://travis-ci.org/ellessenne/comorbidity)
[![Coverage Status](https://img.shields.io/codecov/c/github/ellessenne/comorbidity/master.svg)](https://codecov.io/github/ellessenne/comorbidity?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/comorbidity)](https://cran.r-project.org/package=comorbidity)

`comorbidity` is an R package for computing comorbidity scores such as the weighted Charlson score and the Elixhauser comorbidity score, assuming ICD-10 diagnostic codes.

## Installation

You can install `comorbidity` from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ellessenne/comorbidity")
```

## Example

This is a basic example which shows you how to solve a common problem.

First, we simulate 150 ICD-10 diagnostic codes for 10 individuals using the `sample_codes_icd10()` function:

```{r simulate-data}
# load the comorbidity package
library(comorbidity)
# set a seed for reproducibility
set.seed(1)
x <- data.frame(
  id = sample(1:10, size = 150, replace = TRUE),
  code = sample_diag_icd10(150),
  stringsAsFactors = FALSE)
head(x)
```

Then, we compute the Charlson score, index, and each comorbidity domain:

```{r charlson}
charlson = comorbidity(x = x, id = "id", code = "code", score = "charlson_icd10")
head(charlson)
```

Analogously, using the Elixhauser score:

```{r elixhauser}
elixhauser = comorbidity(x = x, id = "id", code = "code", score = "elixhauser_icd10")
head(elixhauser)
```

## References

This package is based on the ICD-10-based formulations of the Charlson score and Elixhauser score proposed by Quan _et al_. in 2005. Weights for the Charlson score are based on the original formulation by Charlson _et al_. in 1987, while weights for the Elixhauser score are based on work by van Walraven _et al_. Finally, the categorisation of scores and weighted scores is based on work by Menendez _et al_.
 
* Quan H, Sundararajan V, Halfon P, Fong A, Burnand B, Luthi J-C, et al. _Coding algorithms for defining comorbidities in ICD-9-CM and ICD-10 administrative data_. Medical Care 2005; 43(11):1130-1139.
* Charlson ME, Pompei P, Ales KL, et al. _A new method of classifying prognostic comorbidity in longitudinal studies: development and validation_. Journal of Chronic Diseases 1987; 40:373-383.
* van Walraven C, Austin PC, Jennings A, Quan H and Forster AJ. _A modification of the Elixhauser comorbidity measures into a point system for hospital death using administrative data_. Medical Care 2009; 47(6):626-633.
* Menendez ME, Neuhaus V, van Dijk CN, Ring D. _The Elixhauser comorbidity method outperforms the Charlson index in predicting inpatient death after orthopaedic surgery_. Clinical Orthopaedics and Related Research 2014; 472:2878–2886.
