---
title: "Comorbidity scores"
author: "Alessandro Gasparini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comorbidity scores}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center", fig.height = 6, fig.width = 6,
  out.width = "66.66%"
)
```

`comorbidity` is an R package for computing comorbidity scores based on ICD codes data. As of version `0.1.0`, `comorbidity` can calculate the Charlson comorbidity score and the Elixhauser comorbidity score, using either the ICD-9 or ICD-10 coding system. Details on each score supported by this package are presented in this vignette, along with examples on how to compute the different scores with this software.

# Charlson comorbidity score

The Charlson comorbidity index was first developed by Charlson _et al_. in 1987 to predict one-year mortality for patients who may have a range of comorbid conditions. Each condition is assigned a score depending on the risk of dying associated with each one, and consequently scores are summed to provide a total score to predict mortality.

The Charlson comorbidity index includes the following comorbid conditions: acute myocardial infarction, congestive heart failure, peripheral vascular disease, cerebrovascular disease, dementia, chronic obstructive pulmonary disease [COPD], rheumatoid disease, peptic ulcer disease, mild and moderate/severe liver disease, diabetes mellitus with and without complications, hemiplegia/paraplegia, renal disease, cancer (any malignancy) and metastatic solid tumour, AIDS/HIV.

Many variations of the Charlson comorbidity index have been presented, as outlined by Sharabiani _et al_. in their systematic review. `comorbidity` computes the Quan _et al_. version of the Charlson score for both ICD-9-CM and ICD-10 coding systems, as outlined in their paper from 2005; in the next subsections, we present the different ICD codes utilised by `comorbidity`. Categorisation of scores and weighted scores are based on work by Menendez _et al_.

## ICD-9-CM codes

The ICD-9-CM codes used by `comorbidity` to compute the Charlson comorbidity index are:

* Myocardial infarction: 410.x, 412.x

* Congestive heart failure: 398.91, 402.01, 402.11, 402.91, 404.01, 404.03, 404.11, 404.13, 404.91, 404.93, 425.4 - 425.9, 428.x

* Peripheral vascular disease: 093.0, 437.3, 440.x, 441.x, 443.1 - 443.9, 47.1, 557.1, 557.9, V43.4

* Cerebrovascular disease: 362.34, 430.x - 438.x

* Dementia: 290.x, 294.1, 331.2

* Chronic pulmonary disease: 416.8, 416.9, 490.x - 505.x, 506.4, 508.1, 508.8

* Rheumatic disease: 446.5, 710.0 - 710.4, 714.0 - 714.2, 714.8, 725.x

* Peptic ulcer disease: 531.x - 534.x

* Mild liver disease: 070.22, 070.23, 070.32, 070.33, 070.44, 070.54, 070.6, 070.9, 570.x, 571.x, 573.3, 573.4, 573.8, 573.9, V42.7

* Diabetes without chronic complication: 250.0 - 250.3, 250.8, 250.9

* Diabetes with chronic complication: 250.4 - 250.7

* Hemiplegia or paraplegia: 334.1, 342.x, 343.x, 344.0 - 344.6, 344.9

* Renal disease: 403.01, 403.11, 403.91, 404.02, 404.03, 404.12, 404.13, 404.92, 404.93, 582.x, 583.0 - 583.7, 585.x, 586.x, 588.0, V42.0, V45.1, V56.x

* Any malignancy, including lymphoma and leukaemia, except malignant neoplasm of skin: 140.x - 172.x, 174.x - 195.8, 200.x - 208.x, 238.6

* Moderate or severe liver disease: 456.0 - 456.2, 572.2- 572.8

* Metastatic solid tumour: 196.x - 199.x

* AIDS/HIV: 042.x - 044.x

## ICD-10 codes

The ICD-10 codes used by `comorbidity` to compute the Charlson comorbidity index are:

* Myocardial infarction: I21.x, I22.x, I25.2

* Congestive heart failure: I09.9, I11.0, I13.0, I13.2, I25.5, I42.0, I42.5 - I42.9, I43.x, I50.x, P29.0

* Peripheral vascular disease: I70.x, I71.x, I73.1, I73.8, I73.9, I77.1, I79.0, I79.2, K55.1, K55.8, K55.9, Z95.8, Z95.9

* Cerebrovascular disease: G45.x, G46.x, H34.0, I60.x - I69.x

* Dementia: F00.x - F03.x, F05.1, G30.x, G31.1

* Chronic pulmonary disease: I27.8, I27.9, J40.x - J47.x, J60.x - J67.x, J68.4, J70.1, J70.3

* Rheumatic disease: M05.x, M06.x, M31.5, M32.x - M34.x, M35.1, M35.3, M36.0

* Peptic ulcer disease: K25.x - K28.x

* Mild liver disease: B18.x, K70.0 - K70.3, K70.9, K71.3 - K71.5, K71.7, K73.x, K74.x, K76.0, K76.2 - K76.4, K76.8, K76.9, Z94.4

* Diabetes without chronic complication: E10.0, E10.1, E10.6, E10.8, E10.9, E11.0, E11.1, E11.6, E11.8,  E11.9, E12.0, E12.1, E12.6, E12.8,  E12.9, E13.0, E13.1, E13.6, E13.8,  E13.9, E14.0, E14.1, E14.6, E14.8, E14.9

* Diabetes with chronic complication: E10.2 - E10.5, E10.7, E11.2 - E11.5, E11.7, E12.2 - E12.5, E12.7, E13.2 - E13.5, E13.7, E14.2 - E14.5, E14.7

* Hemiplegia or paraplegia: G04.1, G11.4, G80.1, G80.2, G81.x, G82.x, G83.0 - G83.4, G83.9

* Renal disease: I12.0, I13.1, N03.2 - N03.7, N05.2 - N05.7, N18.x, N19.x, N25.0, Z49.0 - Z49.2, Z94.0, Z99.2

* Any malignancy, including lymphoma and leukaemia, except malignant neoplasm of skin: C00.x - C26.x, C30.x - C34.x, C37.x - C41.x, C43.x, C45.x - C58.x, C60.x - C76.x, C81.x - C85.x, C88.x, C90.x - C97.x

* Moderate or severe liver disease: I85.0, I85.9, I86.4, I98.2, K70.4, K71.1, K72.1, K72.9, K76.5, K76.6, K76.7

* Metastatic solid tumour: C77.x - C80.x

* AIDS/HIV: B20.x - B22.x, B24.x

## Weights

Each condition from the Charlson score is assigned a score when computing the weighted Charlson index, irrespectively of the coding system utilised. In particular, diabetes with complications, hemiplegia/paraplegia, renal disease, and malignancies are assigned a score of 2; moderate/severe liver disease is assigned a score of 3; metastatic solid tumour and AIDS/HIV are assigned a score of 6; the remaining comorbidities are assigned a score of 1. `comorbidity` allows the option of applying a hierarchy of comorbidities should a more severe version be present: by choosing to do so (and that is the default behaviour of `comorbidity`) a type of comorbidity is never computed more than once for a given patient.

# Elixhauser comorbidity score

The Elixhauser comorbidity index, analogously as the Charlson comorbidity index, is a method for measuring patient comorbidity based on ICD-9-CM and ICD-10 diagnosis codes found in administrative data developed by Elixhauser _et al_. in 1998. Over time, there have been changes to the Index based on different research. For instance:

* the original index was developed with 30 categories. Now there are variants with 31 categories (Garland _et al_., 2012);
* the list of specific ICD diagnosis codes that are used to identify different categories of comorbidity have been modified and updated from ICD-9-CM to work with ICD-10 coding (Quan _et al_., 2005);
* weighting algorithms were developed, based on the association between comorbidity and death, in order to produce an overall score for the Elixhauser index (van Walraven _et al_., 2009; Moore _et al_., 2016).

`comorbidity` is using the coding definition of Quan _et al_. (2005) for both ICD-9-CM and ICD-10 coding systems; the actual codes and weights utilised by `comorbidity` are introduced in the next subsections.
However, there is no consensus regarding the weighting algorithm, with several competing definitions.
`comorbidity` implements (and returns) both the AHRQ version of the Elixhauser index (Moore _et al_., 2017) and the van Walraven _et al._ (2009) version.
The AHRQ Elixhauser comorbidity score only includes 29 comorbidities; the missing comorbidities are therefore assigned a weight of zero. 
Finally, the categorisation of scores and weighted scores is based on work by Menendez _et al_.

## ICD-9-CM codes

The ICD-9-CM codes used by `comorbidity` to compute the Elixhauser comorbidity index are:

* Congestive heart failure: 398.91, 402.01, 402.11, 402.91, 404.01, 404.03, 404.11, 404.13, 404.91, 404.93, 425.4 - 425.9, 428.x

* Cardiac arrhythmias: 426.0, 426.13, 426.7, 426.9, 426.10, 426.12, 427.0 - 427.4, 427.6 - 427.9, 785.0, 996.01, 996.04, V45.0, V53.3

* Valvular disease: 093.2, 394.x - 397.x, 424.x, 746.3 - 746.6, V42.2, V43.3

* Pulmonary circulation disorders: 415.0, 415.1, 416.x, 417.0, 417.8, 417.9

* Peripheral vascular disorders: 093.0, 437.3, 440.x, 441.x, 443.1 - 443.9, 447.1, 557.1, 557.9, V43.4

* Hypertension, uncomplicated: 401.x

* Hypertension, complicated: 402.x - 405.x

* Paralysis: 334.1, 342.x, 343.x, 344.0 - 344.6, 344.9

* Other neurological disorders: 331.9, 332.0, 332.1, 333.4, 333.5, 333.92, 334.x - 335.x, 336.2, 340.x, 341.x, 345.x, 348.1, 348.3, 780.3, 784.3

* Chronic pulmonary disease: 416.8, 416.9, 490.x - 505.x, 506.4, 508.1, 508.8

* Diabetes, uncomplicated: 250.0 - 250.3

* Diabetes, complicated: 250.4 - 250.9

* Hypothyroidism: 240.9, 243.x, 244.x, 246.1, 246.8

* Renal failure: 403.01, 403.11, 403.91, 404.02, 404.03, 404.12, 404.13, 404.92, 404.93, 585.x, 586.x, 588.0, V42.0, V45.1, V56.x

* Liver disease: 070.22, 070.23, 070.32, 070.33, 070.44, 070.54, 070.6, 070.9, 456.0 - 456.2, 570.x, 571.x, 572.2 - 572.8, 573.3, 573.4, 573.8, 573.9, V42.7

* Peptic ulcer disease, excluding bleeding: 531.7, 531.9, 532.7, 532.9, 533.7, 533.9, 534.7, 534.9 

* AIDS/HIV: 042.x - 044.x

* Lymphoma: 200.x - 202.x, 203.0, 238.6

* Metastatic cancer: 196.x - 199.x

* Solid tumour without metastasis: 140.x - 172.x, 174.x - 195.x

* Rheumatoid arthritis/collagen vascular diseases: 446.x, 701.0, 710.0 - 710.4,  710.8, 710.9, 711.2,  714.x, 719.3, 720.x,  725.x, 728.5, 728.89, 729.30

* Coagulopathy: 286.x, 287.1, 287.3 - 287.5

* Obesity: 278.0

* Weight loss: 260.x - 263.x, 783.2, 799.4

* Fluid and electrolyte disorders: 253.6, 276.x

* Blood loss anaemia: 280.0

* Deficiency anaemia: 280.1 - 280.9, 281.x

* Alcohol abuse: 265.2, 291.1 - 291.3, 291.5 - 291.9, 303.0, 303.9,  305.0, 357.5, 425.5,  535.3, 571.0 - 571.3, 980.x, V11.3

* Drug abuse: 292.x, 304.x, 305.2 - 305.9, V65.42

* Psychoses: 293.8, 295.x, 296.04, 296.14, 296.44, 296.54, 297.x, 298.x

* Depression: 296.2, 296.3, 296.5, 300.4, 309.x, 311

## ICD-10 codes

The ICD-10 codes used by `comorbidity` to compute the Elixhauser comorbidity index are:

* Congestive heart failure: I09.9, I11.0, I13.0, I13.2, I25.5, I42.0, I42.5 - I42.9, I43.x, I50.x, P29.0

* Cardiac arrhythmias: I44.1 - I44.3, I45.6, I45.9, I47.x - I49.x, R00.0, R00.1, R00.8, T82.1, Z45.0, Z95.0

* Valvular disease: A52.0, I05.x - I08.x, I09.1, I09.8, I34.x - I39.x, Q23.0 - Q23.3, Z95.2 - Z95.4

* Pulmonary circulation disorders: I26.x, I27.x, I28.0, I28.8, I28.9

* Peripheral vascular disorders: I70.x, I71.x, I73.1, I73.8, I73.9, I77.1, I79.0, I79.2, K55.1, K55.8, K55.9, Z95.8, Z95.9

* Hypertension, uncomplicated: I10.x

* Hypertension, complicated: I11.x - I13.x, I15.x

* Paralysis: G04.1, G11.4, G80.1, G80.2, G81.x, G82.x, G83.0 - G83.4, G83.9

* Other neurological disorders: G10.x - G13.x, G20.x - G22.x, G25.4, G25.5, G31.2, G31.8, G31.9, G32.x, G35.x - G37.x, G40.x, G41.x, G93.1, G93.4, R47.0, R56.x

* Chronic pulmonary disease: I27.8, I27.9, J40.x - J47.x, J60.x - J67.x, J68.4, J70.1, J70.3

* Diabetes, uncomplicated: E10.0, E10.1, E10.9, E11.0, E11.1, E11.9, E12.0, E12.1, E12.9, E13.0, E13.1, E13.9, E14.0, E14.1, E14.9

* Diabetes, complicated: E10.2 - E10.8, E11.2 - E11.8, E12.2 - E12.8, E13.2 - E13.8, E14.2 - E14.8

* Hypothyroidism: E00.x - E03.x, E89.0

* Renal failure: I12.0, I13.1, N18.x, N19.x, N25.0, Z49.0 - Z49.2, Z94.0, Z99.2

* Liver disease: B18.x, I85.x, I86.4, I98.2, K70.x, K71.1, K71.3 - K71.5, K71.7, K72.x - K74.x, K76.0, K76.2 - K76.9, Z94.4

* Peptic ulcer disease, excluding bleeding: K25.7, K25.9, K26.7, K26.9, K27.7, K27.9, K28.7, K28.9

* AIDS/HIV: B20.x - B22.x, B24.x

* Lymphoma: C81.x - C85.x, C88.x, C96.x, C90.0, C90.2

* Metastatic cancer: C77.x - C80.x

* Solid tumour without metastasis: C00.x - C26.x, C30.x - C34.x, C37.x - C41.x, C43.x, C45.x - C58.x, C60.x - C76.x, C97.x

* Rheumatoid arthritis/collagen vascular diseases: L94.0, L94.1, L94.3, M05.x, M06.x, M08.x, M12.0, M12.3, M30.x, M31.0 - M31.3, M32.x - M35.x, M45.x, M46.1, M46.8, M46.9

* Coagulopathy: D65 - D68.x, D69.1, D69.3 - D69.6

* Obesity: E66.x

* Weight loss: E40.x - E46.x, R63.4, R64

* Fluid and electrolyte disorders: E22.2, E86.x, E87.x

* Blood loss anaemia: D50.0

* Deficiency anaemia: D50.8, D50.9, D51.x - D53.x

* Alcohol abuse: F10, E52, G62.1, I42.6, K29.2, K70.0, K70.3, K70.9, T51.x, Z50.2, Z71.4, Z72.1

* Drug abuse: F11.x - F16.x, F18.x, F19.x, Z71.5, Z72.2

* Psychoses: F20.x, F22.x - F25.x, F28.x, F29.x, F30.2, F31.2, F31.5

* Depression: F20.4, F31.3 - F31.5, F32.x, F33.x, F34.1, F41.2, F43.2

## Weights

The weights for the Elixhauser comorbidity index are included in the following table, depending on the algorithm used for the weighting process:

                                  Comorbidity Domain     AHRQ Algorithm    van Walraven Algorithm
------------------------------------------------------  ----------------  ------------------------
Congestive heart failure                                      9                         7
Cardiac arrhythmias                                          (0)                        5
Valvular disease                                              0                        -1
Pulmonary circulation disorders                               6                         4
Peripheral vascular disorders                                 3                         2
Hypertension (combined uncomplicated and complicated)        -1                         0
Paralysis                                                     5                         7
Other neurological disorders                                  5                         6
Chronic pulmonary disease                                     3                         3
Diabetes, uncomplicated                                       0                         0
Diabetes, complicated                                        -3                         0
Hypothyroidism                                                4                         0
Renal failure                                                 6                         5
Liver disease                                                 4                         11
Peptic ulcer disease, excluding bleeding                      0                         0
AIDS/HIV                                                      0                         0
Lymphoma                                                      6                         9
Metastatic cancer                                            14                        12
Solid tumour without metastasis                               7                         4
Rheumatoid arthritis/collagen vascular diseases               0                         0
Coagulopathy                                                 11                         3
Obesity                                                      -5                        -4
Weight loss                                                   9                         6
Fluid and electrolyte disorders                              11                         5
Blood loss anaemia                                           -3                        -2
Deficiency anaemia                                           -2                        -2
Alcohol abuse                                                -1                         0
Drug abuse                                                   -7                        -7
Psychoses                                                    -5                         0
Depression                                                   -5                        -3

The AHRQ algorithm does not include cardiac arrhythmias, hence a weight of 0 is assigned.

# Examples

The first step consists in loading the `comorbidity` package:

```{r load-package}
library(comorbidity)
```

We can utilise the built-in `sample_diag()` function to simulate ICD diagnostic codes. Both ICD-9 and ICD-10 codes are supported:

```{r simulate-data}
data9 <- data.frame(
  id = sample(1:10, size = 250, replace = TRUE),
  code = sample_diag(n = 250, version = "ICD9_2015"),
  stringsAsFactors = FALSE
)
data9 <- data9[order(data9$id), ]
data10 <- data.frame(
  id = sample(1:10, size = 250, replace = TRUE),
  code = sample_diag(n = 250, version = "ICD10_2011"),
  stringsAsFactors = FALSE
)
data10 <- data10[order(data10$id), ]
```

Then, we can go ahead and compute various comorbidity scores and indices supported by `comorbidity`. The Charlson score based on ICD-9-CM data is computed as:

```{r charlson-9}
charlson9 <- comorbidity(x = data9, id = "id", code = "code", score = "charlson", icd = "icd9", assign0 = FALSE)
str(charlson9)
```

The Charlson score based on ICD-10 data:

```{r charlson-10}
charlson10 <- comorbidity(x = data10, id = "id", code = "code", score = "charlson", icd = "icd10", assign0 = FALSE)
str(charlson10)
```

The Elixhauser score based on ICD-9-CM data:

```{r elixhauser-9}
elixhauser9 <- comorbidity(x = data9, id = "id", code = "code", score = "elixhauser", icd = "icd9", assign0 = FALSE)
str(elixhauser9)
```

Finally, the Elixhauser score based on ICD-10 data:

```{r elixhauser-10}
elixhauser10 <- comorbidity(x = data10, id = "id", code = "code", score = "elixhauser", icd = "icd10", assign0 = FALSE)
str(elixhauser10)
```

# References

* Charlson ME, Pompei P, Ales KL, et al. _A new method of classifying prognostic comorbidity in longitudinal studies: development and validation_. Journal of Chronic Diseases 1987; 40:373-383. DOI: [10.1016/0021-9681(87)90171-8](https://doi.org/10.1016/0021-9681(87)90171-8)

* Sharabiani MT, Aylin P, Bottle A. _Systematic review of comorbidity indices for administrative data_. Medical Care 2012; 50(12):1109-1118. DOI: [10.1097/MLR.0b013e31825f64d0](https://doi.org/10.1097/MLR.0b013e31825f64d0)

* Quan H, Sundararajan V, Halfon P, Fong A, Burnand B, Luthi JC, et al. _Coding algorithms for defining comorbidities in ICD-9-CM and ICD-10 administrative data_. Medical Care 2005; 43(11):1130-1139. DOI: [10.1097/01.mlr.0000182534.19832.83](https://doi.org/10.1097/01.mlr.0000182534.19832.83)

* Menendez ME, Neuhaus V, van Dijk CN, Ring D. _The Elixhauser comorbidity method outperforms the Charlson index in predicting inpatient death after orthopaedic surgery_. Clinical Orthopaedics and Related Research 2014; 472(9):2878-2886. DOI: [10.1007/s11999-014-3686-7](https://doi.org/10.1007/s11999-014-3686-7)

* Elixhauser A, Steiner C, Harris DR and Coffey RM. _Comorbidity measures for use with administrative data_. Medical Care 1998; 36(1):8-27. DOI: [10.1097/00005650-199801000-00004 ](https://doi.org/10.1097/00005650-199801000-00004 )

* Garland A, Fransoo R, Olafson K, Ramsey C, Yogendran M, Chateau D, McGowan K. _The epidemiology and outcomes of critical illness in Manitoba_. Winnipeg, MB: Manitoba Centre for Health Policy, 2012. URL: http://mchp-appserv.cpe.umanitoba.ca/reference/MCHP_ICU_Report_WEB_(20120403).pdf

* van Walraven C, Austin PC, Jennings A, Quan H and Forster AJ. _A modification of the Elixhauser comorbidity measures into a point system for hospital death using administrative data_. Medical Care 2009; 47(6):626-633. DOI: [10.1097/MLR.0b013e31819432e5](https://doi.org/10.1097/MLR.0b013e31819432e5)

* Moore BJ, White S, Washington R, Coenen N, and Elixhauser A. _Identifying increased risk of readmission and in-hospital mortality using hospital administrative data: the AHRQ Elixhauser comorbidity index_. Medical Care 2017; 55(7):698-705. DOI: [10.1097/MLR.0000000000000735](https://doi.org/10.1097/MLR.0000000000000735)
